---
title: "Project_1_data_cookbook"
author: "P. Shao"
date: "November 14, 2015"
output: html_document
---
# Loading Data

```{r,eval=FALSE,tidy=TRUE}
news_raw <- read.table("./News_train.txt", header = FALSE)
hill <- read.table("./Hill-Valley_train.txt", header = FALSE)
news_names <- read.table("./names.txt", header = FALSE, skip = 44, nrows = 59, sep = ":", col.names = c("Name", "Description"))
```

# Basic transformation

1. take the log of V59, saving into `news_reg` named `log_res`, and transform the V59 into factor varible, saving into `news_cls` named `class_res`, where <=1100 is low, >1100 and <=2100 is med, >2100 is high;
2. combine V12 to V17, create a factor variable called `type`, with 7 levels;
3. combine V30 to V36, create a factor variable called `weekday`, with 7 levels;
4. Transform V37 into factor;
5. Drop the columns: 12:17, 30:36, 59;
6. change the variables names.

```{r, eval=FALSE, tidy=FALSE}
library(dplyr)
types <- c(sapply(strsplit(news_names[12:17, 1], "_"), "[", 4), "others")
weekdays <- sapply(strsplit(news_names[30:36, 1], "_"), "[", 3)
drops <- c(12:17, 30:36, 59)

news_reg <- news_raw %>%
        mutate(log_res = log(V59), 
               type = factor(cbind(V12, V13, V14, V15, V16, V17, 
                                   !(V12|V13|V14|V15|V16|V17)) %*% 
                                     matrix(1:length(types)), 
                             levels = 1:length(types), 
                             labels = types), 
               weekday = factor(cbind(V30, V31, V32, V33, V34, V35, V36) %*% 
                                        matrix(1:length(weekdays)), 
                                levels = 1:length(weekdays), 
                                labels = weekdays), 
               V37 = factor(V37, levels = c(1, 0), labels = c(TRUE, FALSE))) %>%
        select(-drops)
names(news_reg)[1:45] <- sapply(strsplit(news_names[, 1], ". "), "[", 2)[-drops]

news_cls <- news_raw %>%
        mutate(class_res = factor(cbind(V59<=1100, V59>1100&V59<=2100, V59>2100) %*%
                                          c(1, 2, 3), 
                                  levels = 1:3, 
                                  labels = c("low", "med", "high")), 
               type = factor(cbind(V12, V13, V14, V15, V16, V17, 
                                   !(V12|V13|V14|V15|V16|V17)) %*% 
                                     matrix(1:length(types)), 
                             levels = 1:length(types), 
                             labels = types), 
               weekday = factor(cbind(V30, V31, V32, V33, V34, V35, V36) %*% 
                                        matrix(1:length(weekdays)), 
                                levels = 1:length(weekdays), 
                                labels = weekdays), 
               V37 = factor(V37, levels = c(1, 0), labels = c(TRUE, FALSE))) %>%
        select(-drops)
names(news_cls)[1:45] <- sapply(strsplit(news_names[, 1], ". "), "[", 2)[-drops]
```

# Tidying Data (Missing Values)
## Strategies:
### 1. No more actions
```{r, eval=FALSE}
news_reg1 <- news_reg
news_cls1 <- news_cls
```

### 2. Remove the data with missing values(the number of words with content is zero)

```{r, eval=FALSE}
news_reg2 <- news_reg
news_reg2 <- news_reg2[!(news_reg2$n_tokens_content == 0), ]
news_cls2 <- news_cls
news_cls2 <- news_cls2[!(news_cls2$n_tokens_content == 0), ]
```

### 3. Separate the data into two parts, zero part and non-zero part

```{r, eval=FALSE}
news_reg3 <- news_reg
news_cls3 <- news_cls
news_reg3_zero <- news_reg3[(news_reg3$n_tokens_content == 0), ]
news_reg3_nonzero <- news_reg3[!(news_reg3$n_tokens_content == 0), ]
news_cls3_zero <- news_cls3[(news_cls3$n_tokens_content == 0), ]
news_cls3_nonzero <- news_cls3[!(news_cls3$n_tokens_content == 0), ]
```


### 4. Let the missing values be average

```{r, eval=FALSE, tidy=FALSE}
news_reg4 <- news_reg
news_cls4 <- news_cls
ind <- (news_reg4$n_tokens_content == 0)
news_reg4[ind, 2] <- round(mean(news_reg4[!ind, 2]))
news_reg4[ind, 3] <- mean(news_reg4[!ind, 3])
news_reg4[ind, 4] <- mean(news_reg4[!ind, 4])
news_reg4[ind, 5] <- mean(news_reg4[!ind, 5])
news_cls4[ind, 2] <- round(mean(news_cls4[!ind, 2]))
news_cls4[ind, 3] <- mean(news_cls4[!ind, 3])
news_cls4[ind, 4] <- mean(news_cls4[!ind, 4])
news_cls4[ind, 5] <- mean(news_cls4[!ind, 5])
```






# Reference
<https://rstudio-pubs-static.s3.amazonaws.com/122671_778c16d46da6489c9f88cd7c12b20ed3.html>